HyperLogLog: Exponentially Bad in Adversarial Settings

### 2020-05-15 docs/HLL_exp_bad/main.tex
authors
fix up

\documentclass{IEEEtran}
\usepackage{graphicx}
\usepackage{array}
\usepackage{url}
\usepackage{hyperref}
\usepackage{xspace}

@@

\title{HyperLogLog: Exponentially Bad in Adversarial Settings}
\author{Mathilde Raynal, Kenneth G. Paterson}

@@

\IEEEtitleabstractindextext{%
\begin{abstract}
Computing the count of distinct elements in massive data sets is a common task but naive approaches are memory-expensive. The HyperLogLog (HLL) algorithm (Flajolet \emph{et al.}, 2007) estimates a data stream's cardinality while using significantly less memory than a naive approach at the cost of some accuracy. This trade-off makes the HLL algorithm very attractive for a wide range of applications such as database management and network monitoring, where an exact count may not be needed. Recently, the HLL algorithm has started to be proposed for use in scenarios where the inputs may be adversarially generated, for example detection of network scanning attacks. This prompts an examination of the performance of the HLL algorithm in the face of adversarial inputs. In this cautionary note, we show that in such a setting, the HLL algorithm's estimate of cardinality can be exponentially bad: when an adversary has access to the internals of the HLL algorithm and has some flexibility in choosing what inputs will be recorded, it can manipulate the cardinality estimate to be exponentially smaller than the true cardinality.
\end{abstract}
}

\IEEEdisplaynontitleabstractindextext

@@

\begin{itemize}
\item[S1:] The adversary has access to an empty black-box replica of the sketch in use. He can use the API to insert elements into that shadow device, get the cardinality estimate, and reset the sketch. The attacker does not know the elements previously added by honest users nor the state. That is we can test and find elements that are unlikely to increase the HLL estimate on the shadow HLL and then insert the on the target HLL but we do not know the state of the target HLL. It would also be reasonable for the provider to transform the items/hash with  a salt e.g. with a different salt per customer. So one can have both an open source implementation and not know the hash.
\item[S2:] The adversary has access to the details of the HLL implementation, i.e., the adversary knows the number of buckets $2^n$ and the hash function $h$ in use, but is otherwise ``blind''. In particular, an S2 adversary does not have access to the values of the HLL cardinality estimate at any point in its attack. This kind of attack scenario is appropriate in the context of the port scanning attack of~\cite{portscanhll}.
\item[S3:] The adversary has access to the details of the HLL implementation, i.e., the adversary knows the number of buckets $2^n$ and the hash function $h$ in use. Additionally, it can access and interact with the sketch via an API provided by the sketch owner, allowing it to insert items into the HLL sketch and ask for the HLL's cardinality estimate at any point in time.
\item[S4:] The adversary has direct access to the HLL in use and all its internals, i.e., the number of buckets, the hash function $h$, and also the contents of each bucket in the HLL sketch \emph{at one specific point in time}, i.e.\ it is given a \emph{snapshot} of the sketch. It then causes chosen inputs to be inserted into the HLL sketch. An S4 adversary can compute a lower bound on the HLL cardinality estimate based on its snapshot and whatever it causes to be inserted into the HLL sketch during its attack (this is irrespective of whether its attack proceeds with input from other users being added to the HLL sketch in parallel with its attack or not; if it knows that there is no other input, then it can compute the \emph{exact} value of the HLL cardinality estimate at every point in its attack instead of a lower bound).
\end{itemize}

@@

We start by highlighting the fact that S1 seems unrealistic in practice. In a networking application, the authors of~\cite{hllvuln} argue that the assumptions of S1 could be fulfilled if the attacker learns which machine is used to do the monitoring and buys the same device. It is reasonable to think that the provider may use a different salt per device, which would make any effect on the shadow device unexpected on the target and would break the whole attack presented in Section 3.2 of \cite{hllvuln}. We consider that the machines use the same exact internals. Assuming that such setup is possible, we show that it is actually possible for the attacker to learn the internals of the HLL sketch. The API must explicitly tell the customer the error of the sketch, and hence the number of buckets. We show that the attacker can learn the hash of a targeted item using the knowlwdege of S1, and thus end up in the S2 scenario. To learn $h(x)$, the attacker can insert the item $x$ on en empty sketch, and ask for the resulting cardinality, which we denote $c$. By solving:
\begin{eqnarray*}
    \alpha 2^{2n}(\sum_{j=1}^{2^n}2^{-c_j})^{-1}& = & c \\
    \Leftrightarrow log_2(c)-log_2(\alpha m^2 - cm + c) &=& M[i].
\end{eqnarray*}
where  the attacker can learn the value of the updated estimator of the bucket $x$ was mapped to, and recover some bits of $h(x)$. This may be possible under certain circumstances, but we do not consider it typical of HLL deployments, and furthermore, it is easy for the attacker to escalate to the knowledge of S2.

Furthermore, we show that although~\cite{cardestprivacy} presents S3 (external adversary) and the stronger version of our S4 (insider adversary) as two separate scenarios, an S3 adversary can easily infer the knowledge given to our S4 adversary in its snapshot. Indeed, an adversary can easily recover the contents of each bucket in the sketch simply by analysing which inputs increase the cardinality estimate. To find the estimate held in a targeted bucket, the attacker gradually inserts items $x$ with an incrementing number of leading zeros in the $\ell$ rightmost bits of $h(x)$ whilst holding the $n$ bits determining the bucket index constant, until the cardinality estimate of the sketch increases. Once a change occurs, the attacker knows that the maximum number of leading zeros observed in that bucket was just updated to a value that can be recovered by looking at the last inserted item. This assumes the attacker has sufficient flexibility in its choice of inputs $x$ so that the relevant bit conditions on $h(x)$ can be forced. The attacker can determine all bucket contents with, in the worst case, $O(2^{n + \ell})$ pre-computation and $O(\ell \cdot 2^{n})$ insert and cardinality estimate queries, and needs about $n + \ell$ bits of flexibility assuming $h$ behaves like a random function. The pre-computation and query complexity required can be much less if the HLL sketch is only moderately loaded.

@@

\noindent\paragraph{Scenario S2} The adversary knows $h$ so it can adjust fields in the item $x$ it is trying to insert and keep only those for which the $\ell$ rightmost bits of $h(x)$ start with a one bit. In this case, the number of leading zero bits computed by the estimator of the destination bucket is always zero, so the estimator of this bucket is unlikely to be updated and the averaged approximate cardinality is likely to be left unchanged. For example, in the scenario of \cite{portscanhll}, the adversary could change the source port of its IP packet until it is satisfied with its hash value. Given sufficient flexibility in the input $x$, the adversary can insert many items into the HLL without increasing the cardinality markedly, by repeatedly sampling from the input domain.

>>>>

\documentclass[11pt]{article}
\usepackage{graphicx}
\usepackage{array}
\usepackage{url}
\usepackage{hyperref}
\usepackage{xspace}
\usepackage{a4wide}

@@

\title{HyperLogLog: Exponentially Bad in Adversarial Settings}
\author{Mathilde Raynal, Kenneth G. Paterson\\
\vspace{3mm}
Department of Computer Science, ETH Zurich\\
{\tt mraynal@student.ethz.ch, kenny.paterson@inf.ethz.ch}}

@@

%\IEEEtitleabstractindextext{%
\begin{abstract}
Computing the count of distinct elements in massive data sets is a common task but naive approaches are memory-expensive. The HyperLogLog (HLL) algorithm (Flajolet \emph{et al.}, 2007) estimates a data stream's cardinality while using significantly less memory than a naive approach at the cost of some accuracy. This trade-off makes the HLL algorithm very attractive for a wide range of applications such as database management and network monitoring, where an exact count may not be needed. Recently, the HLL algorithm has started to be proposed for use in scenarios where the inputs may be adversarially generated, for example detection of network scanning attacks. This prompts an examination of the performance of the HLL algorithm in the face of adversarial inputs. In this cautionary note, we show that in such a setting, the HLL algorithm's estimate of cardinality can be exponentially bad: when an adversary has access to the internals of the HLL algorithm and has some flexibility in choosing what inputs will be recorded, it can manipulate the cardinality estimate to be exponentially smaller than the true cardinality.
\end{abstract}
%}

%\IEEEdisplaynontitleabstractindextext

@@

\begin{itemize}
\item[S1:] The adversary does not know the details of the target HLL implementation but can access a copy of the HLL instance as a user. He can use the API to insert elements into this shadow sketch, get its cardinality estimate, and reset it to its native empty state. The attacker does not know the state of the targeted HLL sketch, meaning that he does not have the list of elements previously inserted into it by other users, nor can he obtain its cardinality estimate.
\item[S2:] The adversary has access to the details of the HLL implementation, i.e., the adversary knows the number of buckets $2^n$ and the hash function $h$ in use, but is otherwise ``blind''. In particular, an S2 adversary does not have access to the values of the HLL cardinality estimate at any point in its attack. This kind of attack scenario is appropriate in the context of the port scanning attack of~\cite{portscanhll}.
\item[S3:] The adversary has access to the details of the HLL implementation, i.e., the adversary knows the number of buckets $2^n$ and the hash function $h$ in use. Additionally, it can access and interact with the sketch via an API provided by the sketch owner, allowing it to insert items into the HLL sketch and ask for the HLL's cardinality estimate at any point in time.
\item[S4:] The adversary has direct access to the HLL in use and all its internals, i.e., the number of buckets, the hash function $h$, and also the contents of each bucket in the HLL sketch \emph{at one specific point in time}, i.e.\ it is given a \emph{snapshot} of the sketch. It then causes chosen inputs to be inserted into the HLL sketch. An S4 adversary can compute a lower bound on the HLL cardinality estimate based on its snapshot and whatever it causes to be inserted into the HLL sketch during its attack (this is irrespective of whether its attack proceeds with input from other users being added to the HLL sketch in parallel with its attack or not; if it knows that there is no other input, then it can compute the \emph{exact} value of the HLL cardinality estimate at every point in its attack instead of a lower bound).
\end{itemize}

@@

We start by highlighting the fact that S1 seems unrealistic in practice. In a networking application, the authors of~\cite{hllvuln} argue that the assumptions of S1 could be fulfilled if the attacker learns which machine is used to do the monitoring, buys the same device, and uses it as the shadow device. It is possible that the provider uses a different HLL salt or key per device. This would make any pre-computation with the shadow device irrelevant and considerably weaken the attack presented in Section 3.2 of \cite{hllvuln}. If, on the other hand, we assume that the devices are perfect copies of one another, then we show that it is actually possible for the attacker to predict the behaviour of the HLL sketch as elements are inserted into it, and effectively elevate itself to being an S2 adversary, without actually knowing the hash function $h$ in use. In detail, the attacker can recover some bits of interest in the hash values of targeted items, and infer the knowledge given to the adversary under S2. To do so, the attacker can insert the item $x$ on an empty sketch and ask for the resulting cardinality, denoted $c$. Note that inserting $x$ only touches one bucket of the HLL sketch; let that bucket have index $i$ (which is unknown to the attacker) and value $c_i$ after the insertion.  From the formula used to compute the HLL sketch cardinality estimate $c$, we get the equation: 
\begin{eqnarray*}
   \log_2(c)-\log_2(\alpha 2^{2n} - c2^n + c) &=& c_i
\end{eqnarray*}
which enables the adversary to learn the value $c_i$ from $c$. (Here $\alpha$ is a constant, and $2^n$ is known since the provider must explicitly tell the customer the expected error of the estimation, hence the number of buckets.) Now $c_i-1$ is (an approximation of) the number of leading zeros in the $l$ rightmost bits of $h(x)$. This is precisely the information extracted from $h(x)$ and used by the adversary in his attack in scenario S2, as we detail in Section~\ref{sec:attacks}. %the value of the updated estimator of the bucket $x$ was mapped to, and recover some bits of $h(x)$.
To summarize, we do not consider scenario S1 typical of HLL deployments, and even if such a setup may be possible under certain circumstances, the attacker can infer the information used in our attack in scenario S2.

%\mathilde{Do you think that we should mention the idea behind the attack of \cite{hllvuln} Section 3.2 ? As they explained in their emails "The attacker can then test and find elements that do not increase the HLL estimate on the shadow HLL and thus will unlikely increase the cardinality of the target HLL when inserted."}

Furthermore, we show that although~\cite{cardestprivacy} presents S3 (external adversary) and the stronger version of our S4 (insider adversary) as two separate scenarios, an S3 adversary can easily infer the knowledge given to our S4 adversary in its snapshot. Indeed, an adversary can easily recover the contents of each bucket in the sketch simply by analysing which inputs increase the cardinality estimate. To find the estimate held in a targeted bucket, the attacker gradually inserts items $x$ with an incrementing number of leading zeros in the $\ell$ rightmost bits of $h(x)$ whilst holding the $n$ bits determining the bucket index constant, until the cardinality estimate of the sketch increases. Once an increase occurs, the attacker knows that the maximum number of leading zeros observed in that bucket was just updated to a value that can be recovered by looking at the last inserted item. This assumes the attacker has sufficient flexibility in its choice of inputs $x$ so that the relevant bit conditions on $h(x)$ can be forced. The attacker can determine all bucket contents with, in the worst case, $O(2^{n + \ell})$ pre-computation and $O(\ell \cdot 2^{n})$ insert and cardinality estimate queries, and needs about $n + \ell$ bits of flexibility assuming $h$ behaves like a random function. The pre-computation and query complexity required can be much less if the HLL sketch is only moderately loaded.

@@

\noindent\paragraph{Scenario S2} The adversary knows $h$ so it can adjust fields in the item $x$ it is trying to insert and keep only those for which the $\ell$ rightmost bits of $h(x)$ start with a one bit. In this case, the number of leading zero bits computed by the estimator of the destination bucket is always zero, so the estimator of this bucket is unlikely to be updated and the averaged approximate cardinality is likely to be left unchanged. For example, in the scenario of \cite{portscanhll}, the adversary could change the source port of its IP packet until it is satisfied with its hash value. Given sufficient flexibility in the input $x$, the adversary can insert many items into the HLL sketch without increasing the cardinality markedly, by repeatedly sampling from the input domain.
